---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-",
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 300
)
```

# commaKit

<!-- badges: start -->
[![R-CMD-check](https://github.com/carl-stone/commaKit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/carl-stone/commaKit/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**commaKit** (**Com**parative **M**icrobial **M**ethylomics **A**nalysis
**Kit**) is an R/Bioconductor-style package for comparative analysis of
bacterial DNA methylation from Oxford Nanopore sequencing data.

commaKit starts from per-sample methylation evidence - usually modkit pileup
bedMethyl files - and builds a `commaData` object, an S4 class extending
`RangedSummarizedExperiment`. From that shared container, the package supports
quality control, genomic annotation, differential methylation testing,
region-level summaries, enrichment analysis, export, and visualization for
6mA, 5mC, and 4mC methylation in monoploid microbial genomes.

commaKit is under active development and is **not yet on Bioconductor**. Install
from GitHub for now.

## Why commaKit?

Nanopore methylation callers can report rich modified-base evidence, but most
analysis workflows still leave users to reconcile caller-specific formats,
coverage filters, genome coordinates, methylation contexts, sample metadata, and
downstream statistics by hand. commaKit provides one R-native container and a
coherent workflow for comparative microbial methylomics:

- import modkit, Dorado, or Megalodon output into a single `commaData` object;
- keep raw methylation and count evidence together with genomic ranges and
  sample metadata;
- analyze methylation by modification context, such as `6mA_GATC` or
  `5mC_CCWGG`;
- test for differential methylation with `methylKit`, `limma`, or a native
  quasi-binomial F-test backend;
- annotate methylation sites to genes or other genomic features;
- summarize sites over regions and export BED-compatible tracks;
- visualize coverage, beta distributions, PCA, genome tracks, metagene/TSS
  profiles, volcano plots, and heatmaps.

The package is designed for bacterial and archaeal genomes where methylation is
often motif-associated, strand-aware, and close to fully methylated or
unmethylated at individual sites.

## Installation

```{r install, eval = FALSE}
# Install the development version from GitHub.
# commaKit is not yet on Bioconductor.
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
devtools::install_github("carl-stone/commaKit")
```

## Input contract

commaKit is intentionally strict about the evidence it imports. The most common
path is modkit pileup bedMethyl:

- Use complete modkit pileup bedMethyl files, not generic BED-like tables.
- Preserve tab-delimited bedMethyl structure so blank required fields are not
  shifted into the wrong columns.
- For modkit input, `Nvalid_cov`, `Nmod`, `Ncanonical`, and `Nother_mod` are the
  authoritative count fields. Methylation beta values are computed from
  `Nmod / Nvalid_cov`, not from the derived `fraction_modified` percentage.
- Genome names must match imported methylation chromosomes. If the methylation
  files contain chromosomes absent from `genome`, construction stops with a
  clear error instead of guessing.
- Direct Dorado BAM parsing requires mapped BAM files with MM/ML tags preserved;
  in practice, many users will prefer to run modkit first and import the
  resulting bedMethyl files.

A typical `commaData()` call needs a named vector of per-sample files, a sample
metadata table, genome lengths or a FASTA/BSgenome, and optionally annotation
and motif information:

```{r construct, eval = FALSE}
library(commaKit)

files <- c(
  ctrl_1  = "path/to/ctrl_rep1.bed",
  ctrl_2  = "path/to/ctrl_rep2.bed",
  ctrl_3  = "path/to/ctrl_rep3.bed",
  treat_1 = "path/to/treat_rep1.bed",
  treat_2 = "path/to/treat_rep2.bed",
  treat_3 = "path/to/treat_rep3.bed"
)

col_data <- data.frame(
  sample_name = names(files),
  condition = c("control", "control", "control",
                "treatment", "treatment", "treatment"),
  replicate = c(1L, 2L, 3L, 1L, 2L, 3L)
)

obj <- commaData(
  files = files,
  colData = col_data,
  genome = "path/to/genome.fa",
  annotation = "path/to/genes.gff3",
  motif = "GATC",
  min_coverage = 5L,
  caller = "modkit"
)
```

For import details and common failure modes, see
`vignette("import-troubleshooting", package = "commaKit")`.

## Quick start with example data

commaKit includes a small synthetic dataset for examples and tests. It contains
588 methylation sites across six samples on a 100 kb toy chromosome: 393
6mA/GATC sites and 195 5mC/CCWGG sites.

```{r load-data}
library(commaKit)
data(comma_example_data)
comma_example_data
```

The core accessors expose sample metadata, genomic site metadata, modification
contexts, beta values, coverage, and raw count assays:

```{r inspect-object}
modContexts(comma_example_data)
sampleInfo(comma_example_data)[, c("sample_name", "condition", "replicate")]
dim(methylation(comma_example_data))
dim(siteCoverage(comma_example_data))
```

## Standard workflow

### 1. Quality control

Start by checking per-sample methylation distributions and coverage.

```{r qc-summary}
ms <- methylomeSummary(comma_example_data)
ms[, c(
  "sample_name", "condition", "mean_beta", "median_beta",
  "frac_methylated", "n_covered"
)]
```

```{r plot-coverage, fig.alt = "Coverage depth distribution per sample."}
plot_coverage(comma_example_data)
```

```{r plot-dist, fig.alt = "Beta value density per sample, faceted by modification type."}
plot_methylation_distribution(comma_example_data)
```

`plot_pca()` uses M-values internally rather than raw beta values, which is more
stable for distance-based analyses when many sites are near 0 or 1.

```{r plot-pca, fig.alt = "PCA of methylation profiles colored by condition."}
plot_pca(comma_example_data, color_by = "condition")
```

### 2. Annotate methylation sites

`annotateSites()` maps 1-bp methylation sites to genomic features with
`GenomicRanges::findOverlaps()`. It keeps all associations in list-columns rather
than collapsing each site to a single closest feature.

```{r annotate}
annotated <- annotateSites(comma_example_data, keep = "overlap")

si <- siteInfo(annotated)
mean(lengths(si$feature_names) > 0)
```

```{r plot-metagene, fig.alt = "Mean methylation across gene bodies."}
plot_metagene(comma_example_data, feature = "gene")
```

### 3. Test differential methylation

`diffMethyl()` follows the familiar pattern used by many Bioconductor workflows:
fit the model, then extract and filter a results table. For the current v1 API,
formulas must name one two-level comparison variable, such as `~ condition`.
Multi-factor formulas are intentionally deferred.

```{r diff-methyl}
cd_dm <- diffMethyl(
  comma_example_data,
  formula = ~condition,
  mod_type = "6mA"
)

res <- results(cd_dm)
sig <- filterResults(cd_dm, padj = 0.05, delta_beta = 0.2)

cat("Total 6mA sites tested:", nrow(res), "\n")
cat("Significant sites:", nrow(sig), "\n")

head(res[
  order(res$dm_padj),
  c("chrom", "position", "mod_context", "dm_delta_beta", "dm_padj")
])
```

Effect sizes are reported on the beta scale, and multiple-testing correction is
applied genome-wide across all tested modification contexts in the call.

```{r plot-volcano, fig.alt = "Volcano plot of differential 6mA methylation."}
plot_volcano(res)
```

```{r plot-heatmap, fig.alt = "Heatmap of top differentially methylated 6mA sites."}
plot_heatmap(res, cd_dm, n_sites = 30L)
```

### 4. Summarize regions and export

Use `summarizeRegions()` when site-level methylation should be aggregated over
features, windows, or other genomic intervals.

```{r summarize-regions}
regions <- GenomicRanges::GRanges(
  seqnames = "chr_sim",
  ranges = IRanges::IRanges(start = c(1, 50001), end = c(50000, 100000)),
  region_id = c("left_half", "right_half")
)

region_summary <- summarizeRegions(comma_example_data, regions)
head(region_summary)
```

Export browser-compatible methylation tracks with `writeBED()`:

```{r write-bed, eval = FALSE}
writeBED(cd_dm, file = "results_6mA.bed", sample = "ctrl_1", mod_type = "6mA")
```

### 5. Enrich annotated differential methylation results

`enrichMethylation()` runs GO or KEGG enrichment on genes associated with
differentially methylated sites. The bundled example data use synthetic gene IDs,
so real biological enrichment examples require real bacterial annotations and
identifier mappings.

```{r enrichment, eval = FALSE}
cd_dm <- annotateSites(cd_dm, keep = "overlap")

enr <- enrichMethylation(
  cd_dm,
  ont = "BP",
  gene_role = "target"
)
```

For KEGG workflows, prefer the offline mapping path so routine analyses do not
depend on live API calls:

```{r kegg-offline, eval = FALSE}
id_map <- buildKEGGGeneIDMap(
  "eco",
  OrgDb = org.EcK12.eg.db::org.EcK12.eg.db
)
kegg_t2g <- buildKEGGTermGene("eco", file = "kegg_eco.rds", id_map = id_map)

enr_kegg <- enrichMethylation(
  cd_dm,
  kegg_term2gene = kegg_t2g$term2gene,
  kegg_term2name = kegg_t2g$term2name
)
```

## The `commaData` object

`commaData` extends `RangedSummarizedExperiment`:

- rows are 1-bp genomic methylation sites stored in `rowRanges()`;
- columns are samples stored in `colData()`;
- assays store methylation beta values, coverage, and count evidence;
- site metadata include `mod_type` and `motif`;
- `mod_context` is computed from `mod_type` and `motif`, for example
  `6mA_GATC`; when motif is unavailable, the modification type is used alone;
- named assay and result layers preserve derived transformations and
  differential methylation runs without overwriting raw evidence.

```{r assay-and-result-layers}
assayLayers(comma_example_data)
resultLayers(cd_dm)
```

Raw evidence assays are treated as canonical input evidence. Transformations,
when created, should live as named derived assay layers. Filtering returns subset
`commaData` objects rather than hidden lazy views.

## Multi-modification analyses

A single object can hold multiple modification types and motifs. Most analysis
and plotting functions accept `mod_type`, `motif`, or `mod_context` filters.

```{r mod-types}
modTypes(comma_example_data)
modContexts(comma_example_data)

obj_6mA <- filterSites(comma_example_data, mod_context = "6mA_GATC")
obj_6mA
```

For a complete joint 6mA + 5mC workflow, see
`vignette("multiple-modification-types", package = "commaKit")`.

## Documentation

```{r docs-help, eval = FALSE}
?commaKit
?commaData
?diffMethyl
?results
```

Vignettes:

- `vignette("getting-started", package = "commaKit")` - end-to-end analysis
  from example data.
- `vignette("understanding-commaData", package = "commaKit")` - object model,
  assays, provenance, and result layers.
- `vignette("import-troubleshooting", package = "commaKit")` - modkit, Dorado,
  Megalodon, genome, and annotation input issues.
- `vignette("multiple-modification-types", package = "commaKit")` - joint
  analysis of multiple modification contexts.

## Current status and limitations

| Area | Status |
|---|---|
| Package version | `0.2.0` development baseline |
| Distribution | GitHub only; not yet submitted to Bioconductor |
| Primary input | modkit pileup bedMethyl |
| Optional inputs | Dorado BAM with MM/ML tags; Megalodon legacy output |
| Differential methylation | One two-level design variable for v1 |
| Example data | Synthetic, deterministic, useful for tests and tutorials |
| Real-data validation | Planned; depends on selecting a redistributable bacterial dataset |
| Performance evidence | Planned before broader release confidence |

Roadmap:

| Version | Phase | Status |
|---|---|---|
| 0.2.0 | Schema v2, commaKit rename, result layers, assay provenance | Done |
| 0.2.x | Test quality, parser hardening, docs synchronization | In progress |
| 0.x.y | Real-data examples, performance benchmarks, Bioconductor hardening | Planned |
| 0.99.0 | Bioconductor submission version | Future |
| 1.0.0 | Stable public release after external confidence | Future |

## Support

Use GitHub Issues for bug reports, feature requests, and documentation problems:

<https://github.com/carl-stone/commaKit/issues>

For import problems, include the caller (`modkit`, `dorado`, or `megalodon`), the
relevant command used upstream, a small excerpt of input if possible, and the
full error message.

## Citation

If you use commaKit in published work, please cite the repository for now and the
related methylation study:

Stone CJ et al. (2022). Genome-wide adenine methylation in *Escherichia coli*
K-12 reveals methylation patterns associated with gene regulation. *bioRxiv*.
